---
title: "Yeast Genome Analysis: Loading, Subsetting, and RNA-to-DNA Conversion"
author: "Adegun Oluwafemi Joseph"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

## Introduction

This analysis demonstrates loading and manipulating *Saccharomyces cerevisiae* (baker's yeast) genome data using the `Biostrings` and `BSgenome` packages in R. The workflow includes: data import, subsetting sequences, and RNA-to-DNA conversions.

------------------------------------------------------------------------

## Load Required Libraries

The following code loads the essential bioinformatics packages for genome data manipulation:

```{r load-libraries, message=TRUE}
# Load the Biostrings package for biological sequence manipulation
# Provides functions for working with DNA, RNA, and protein sequences
library(Biostrings)

# Load the pre-compiled S. cerevisiae reference genome (sacCer3 assembly)
# from the BSgenome package ecosystem
library(BSgenome.Scerevisiae.UCSC.sacCer3)
```

------------------------------------------------------------------------

## Load Yeast Genome Data

Assign the *S. cerevisiae* genome object to a variable for downstream analysis:

```{r load-yeast-genome}
# Assign the S. cerevisiae reference genome (sacCer3) to a variable
# This is a BSgenome object containing all 16 nuclear chromosomes
yeast <- BSgenome.Scerevisiae.UCSC.sacCer3

# Verify the object class and structure
class(yeast)

# Display basic genome information
yeast
```

------------------------------------------------------------------------

## Subset Genome Data

Extract specific chromosomes or regions of interest:

```{r subset-chromosome}
# Extract chromosome I (the smallest yeast chromosome)
chr1 <- yeast$chrI

# Display sequence information
class(chr1)
length(chr1)

# Extract a specific region (first 1000 base pairs of chromosome I)
region_subset <- subseq(chr1, start = 1, end = 1000)
region_subset
```

------------------------------------------------------------------------

## RNA-to-DNA Conversion

Convert RNA sequences to DNA and vice versa:

```{r rna-dna-conversion}
# Convert a DNA sequence to RNA (replace T with U)
dna_seq <- DNAString("ATGCGTACG")
rna_from_dna <- RNAString(dna_seq)

cat("Original DNA:", as.character(dna_seq), "\n")
cat("Converted to RNA:", as.character(rna_from_dna), "\n\n")

# Reverse conversion: RNA to DNA (replace U with T)
rna_seq <- RNAString("AUGCGUACG")
dna_from_rna <- DNAString(rna_seq)

cat("Original RNA:", as.character(rna_seq), "\n")
cat("Converted to DNA:", as.character(dna_from_rna), "\n")
```

------------------------------------------------------------------------

## Session Information

```{r session-info}
# Display R version and loaded packages for reproducibility
sessionInfo()
```
